Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS

### 2023-08-11 paper/ms.tex

Data from K2 and TESS have yielded a new class of variable star whose
root cause is only beginning to become clear: complex quasiperiodic
variables (CQVs).  These objects are identified 
from their optical light curves, which show nearly periodic troughs
that can be either sharp or broad, often superposed on smooth
spot-like modulation
\citep{2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z}.
Some CQVs show up to eight local minima (dips) per cycle.  Most are
pre-main-sequence M dwarfs without near-infrared excesses, ages of
$\approx$5-150 million years (Myr), and rotation periods of at most
two days; they are observed to comprise $\approx$1-3\% of M dwarfs younger
than 100 Myr \citep{2016AJ....152..114R,2022AJ....163..144G}.  The
dips can be chromatic, with a reddening law plausibly consistent with
dust
\citep{2020AJ....160...86B,2022AJ....163..144G,2023MNRAS.518.2921K}.
And finally, while the dip shapes can ``jump'' between different
depths and durations over less than one cycle, they more often evolve
gradually, over tens to hundreds of cycles
\citep[e.g.][]{2017AJ....153..152S,2022ApJ...925...75P,2023ApJ...945..114P}.

@@

We analyzed the ``short'' 120-second cadence data acquired by TESS
between July~2018 and Sep~2022 (Sectors 1-55).  Specifically, we used
the 120-second cadence light curve products produced by the 
Science Processing and Operations Center at NASA Ames
\citep{2016SPIE.9913E..3EJ}.  While the TESS data products also
include full frame images with cadences of 200, 600, and 1800 seconds,
we limited our scope in this work for the sake of simplicity in data
handling.  In exchange, we lose in both completeness and homogeneity
of the selection function.  While TESS cumulatively observed
$\approx$90\% of the sky for at least one lunar month between
July~2018 and Sep~2022, the 120-second cadence data were acquired for
only a pre-selected set of stars over Sectors 1-26, and then a
guest-investigator driven set of stars over Sectors 27-55
\citep{2021PASP..133i5002F}.

@@

\subsubsection{Fourier analysis}
\label{subsec:fourier}
For the Fourier analysis, we followed \citet{2019ApJ...876..127Z}.

{\bf TODO for Rahul or Saul: explain the approach, in a few
	paragraphs.  Was the SAP\_FLUX or PDCSAP used? etc. }

@@

Broadly speaking, the most common false positives for both the Fourier
and dip-counting techniques were eclipsing binaries, spot-induced
variability from rapid rotators, and variability from neighboring,
off-target stars.  Typical false positive rates from our dip-counting
pipeline were 5:1, with \nuniqdipflagged\ unique stars flagged, and
about 20\% being labeled either ``good'' or ``possible'' CQVs; for
the Fourier pipeline, {\bf the rate was X:Y, with ZZZZ unique stars
flagged, and NN\% being labeled ``good'' or ``possible'' CQVs}.

@@

\software{
  %\texttt{arviz} \citep{arviz_2019},
  %\texttt{altaipony} \citep{ilin_flares_2021},
  \texttt{astrobase} \citep{2021zndo...1011188B},
  %\texttt{astroplan} \citep{astroplan2018},
	%\texttt{AstroImageJ} \citep{collins_astroimagej_2017},
  %\texttt{astropy} \citep{astropy_2018},
 % \texttt{astroquery} \citep{astroquery_2018},
  %\texttt{BATMAN} \citep{kreidberg_batman_2015},
  %\texttt{ceres} \citep{brahm_2017_ceres},
  %\texttt{cdips-pipeline} \citep{bhatti_cdips-pipeline_2019},
  %\texttt{corner} \citep{corner_2016},
  %\texttt{emcee} \citep{foreman-mackey_emcee_2013},
  %\texttt{exoplanet} \citep{exoplanet:exoplanet}, and its
  %dependencies \citep{exoplanet:agol20, exoplanet:kipping13, exoplanet:luger18,
  % 	exoplanet:theano},
	%\texttt{gala} \citep{gala,PriceWhelan_2017_gala_zenodo},
	%\texttt{IDL Astronomy User's Library} \citep{landsman_1995},
  %\texttt{IPython} \citep{perez_2007},
	%\texttt{isochrones} \citep{morton_2015_isochrones},
	\texttt{lightkurve} \citep{2018ascl.soft12013L},
  %\texttt{matplotlib} \citep{hunter_matplotlib_2007}, 
  %\texttt{MESA} \citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015}
  %\texttt{numpy} \citep{walt_numpy_2011}, 
  %\texttt{pandas} \citep{mckinney-proc-scipy-2010},
  \texttt{pyGAM} \citep{daniel_serven_2018_1208723},
  %\texttt{PyMC3} \citep{salvatier_2016_PyMC3},
  %\texttt{radvel} \citep{fulton_radvel_2018},
  %\texttt{scikit-learn} \citep{scikit-learn},
  \texttt{scipy} \citep{2020NatMe..17..261V},
  \texttt{TESS-point}  \citep{2020ascl.soft03001B},
  %\texttt{tesscut} \citep{brasseur_astrocut_2019},
  %\texttt{unpopular} \citep{hattorio_2021_cpm},
  %\texttt{VESPA} \citep{morton_efficient_2012,vespa_2015},
  %\texttt{webplotdigitzer} \citep{rohatgi_2019},
  \texttt{wotan} \citep{hippke_wotan_2019}.
}
\

>>>>

Data from K2 and TESS have yielded a new class of variable star whose
root cause is only beginning to become clear: complex quasiperiodic
variables (CQVs).  These objects are identified 
from their optical light curves, which show nearly periodic troughs
that are either sharp or broad; these troughs are often superposed on
smooth spot-like modulation
\citep{2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z}.
Some CQVs show up to eight local minima (``dips'') per cycle.  Most are
pre-main-sequence M dwarfs without near-infrared excesses, ages of
$\approx$5-150 million years (Myr), and rotation periods of at most
two days; they are observed to comprise $\approx$1-3\% of M dwarfs younger
than 100 Myr \citep{2016AJ....152..114R,2022AJ....163..144G}.  The
dips can be chromatic, with a reddening law plausibly consistent with
dust
\citep{2020AJ....160...86B,2022AJ....163..144G,2023MNRAS.518.2921K}.
And finally, while the dip shapes can ``jump'' between different
depths and durations over less than one cycle, they more often evolve
gradually, over tens to hundreds of cycles
\citep[e.g.][]{2017AJ....153..152S,2022ApJ...925...75P,2023ApJ...945..114P}.

@@

We analyzed the ``short'' 120-second cadence data acquired by TESS
between July~2018 and Sep~2022 (Sectors 1-55).  Specifically, we used
the 120-second cadence light curves produced by the Science
Processing and Operations Center at NASA Ames
\citep{2016SPIE.9913E..3EJ}.  While the TESS data products from these
sectors also included full frame images with cadences of 600 and 1800
seconds, we limited our scope in this study for the sake of simplicity
in data handling.  In exchange, we sacrificed both completeness and
homogeneity of the selection function.  While TESS cumulatively
observed $\approx$90\% of the sky for at least one lunar month between
July~2018 and Sep~2022, the 120-second cadence data were collected for
only a subset of observable stars during this time due to telemetric
constraints
\citep[see][]{2021PASP..133i5002F}.

@@

\subsubsection{Fourier analysis}
\label{subsec:fourier}

We performed an independent search for strongly periodic phenomena
using our Fourier pipeline, following \citet{2019ApJ...876..127Z} and
\citet[][their Section~1.3]{2023MNRAS.524.4220P}.
Starting with the PDC\_SAP light curves, we normalized each light
curve, and then re-binned it into equal width bins {\bf LB: how long?} to
account for the uneven spacing in the TESS data, as well as the data
gap during each sector caused by satellite downlink.  We then padded
the data to ensure that the light curve had a length that was a power
of two, as described by \citeauthor{2019ApJ...876..127Z}.  We
then took the Fourier transform of the corrected, padded light curve using 
{\tt numpy.fft.fft} and searched for peaks with a significance of over 
12-$\sigma$ within each set of 500 bins.

If a peak of such significance was found, we generated a ``summary
sheet'' with information about the star, its full and folded light
curves, Fourier transform, potential contaminating stars, and
information about these contaminating stars.  
In the sectors for which
we performed this analysis, we found that ``summary'' sheets were
generated for approximately 10\% of the 20\,000 120-second targets in
a given sector.  
We then manually reviewed each summary sheet, and
visually labelled each star based on its morphology (including e.g.,
eclipsing binary, CQV, RS CVn, or cataclysmic variable).  CQVs that
were observed in multiple sectors had their light curves stitched
together and analyzed for a more accurate period determination.

@@

Broadly speaking, the most common false positives for both the Fourier
and dip-counting techniques were eclipsing binaries, spot-induced
variability from rapid rotators, and variability from neighboring,
off-target stars.  Typical false positive rates from our dip-counting
pipeline were 5:1, with \nuniqdipflagged\ unique stars flagged, and
about 20\% being labeled either ``good'' or ``possible'' CQVs.
The Fourier analysis was not amenable to calculating a similar
false-positive probability, because it was implemented as a general
variability search; its results are also being used for other types of
variability analyses beyond CQVs.

%TODO: include?
%Between the targets identified as CQVs by both pipelines,
% {\red 
%\begin{itemize}
%\item XX targets were found to overlap
%\item YY targets were found via Fourier but not via dip counting
%\end{itemize}
%}
%The difference in discovery rate between these techniques can be attributed to
%low-amplitude peaks in the periodogram that were missed, some of which likely
%arose from fainter CQVs.

@@

\software{
  %\texttt{arviz} \citep{arviz_2019},
  %\texttt{altaipony} \citep{ilin_flares_2021},
  \texttt{astrobase} \citep{2021zndo...1011188B},
  %\texttt{astroplan} \citep{astroplan2018},
	%\texttt{AstroImageJ} \citep{collins_astroimagej_2017},
  %\texttt{astropy} \citep{astropy_2018},
 % \texttt{astroquery} \citep{astroquery_2018},
  \texttt{astropy} \citep{astropy_2013,astropy_2018,astropy_2022},
  %\texttt{astroquery} \citep{astroquery_2018},
  %\texttt{BATMAN} \citep{kreidberg_batman_2015},
  %\texttt{ceres} \citep{brahm_2017_ceres},
  %\texttt{cdips-pipeline} \citep{bhatti_cdips-pipeline_2019},
  %\texttt{corner} \citep{corner_2016},
  %\texttt{emcee} \citep{foreman-mackey_emcee_2013},
  %\texttt{exoplanet} \citep{exoplanet:exoplanet}, and its
  %dependencies \citep{exoplanet:agol20, exoplanet:kipping13, exoplanet:luger18,
  % 	exoplanet:theano},
	%\texttt{gala} \citep{gala,PriceWhelan_2017_gala_zenodo},
	%\texttt{IDL Astronomy User's Library} \citep{landsman_1995},
  %\texttt{IPython} \citep{perez_2007},
	%\texttt{isochrones} \citep{morton_2015_isochrones},
	\texttt{lightkurve} \citep{2018ascl.soft12013L},
  %\texttt{matplotlib} \citep{hunter_matplotlib_2007}, 
  %\texttt{MESA} \citep{paxton_modules_2011,paxton_modules_2013,paxton_modules_2015}
  \texttt{numpy} \citep{2020Natur.585..357H}, 
  %\texttt{pandas} \citep{mckinney-proc-scipy-2010},
  \texttt{pyGAM} \citep{daniel_serven_2018_1208723},
  %\texttt{PyMC3} \citep{salvatier_2016_PyMC3},
  %\texttt{radvel} \citep{fulton_radvel_2018},
  %\texttt{scikit-learn} \citep{scikit-learn},
  \texttt{scipy} \citep{2020NatMe..17..261V},
  \texttt{TESS-point}  \citep{2020ascl.soft03001B},
  %\texttt{tesscut} \citep{brasseur_astrocut_2019},
  %\texttt{unpopular} \citep{hattorio_2021_cpm},
  %\texttt{VESPA} \citep{morton_efficient_2012,vespa_2015},
  %\texttt{webplotdigitzer} \citep{rohatgi_2019},
  \texttt{wotan} \citep{hippke_wotan_2019}.
}
\
